Linear analysis of the vertical shear instability in astrophysical disks

### 2015-02-03 appendix.tex
updating

Using the notation and disk models in the main text, the
energy, density and $r$ momentum equations become 
\begin{align}
  \ii \sigma W  &=c _s^2(r_0, z)\left( \left.\frac{\p\ln\rho}{\p
      z}\right|_0 \delta v_z + \frac{\gamma}{\Gamma} \frac{d\delta
    v_z}{dz}\right) + \left[\ii k_rc_s^2(r_0,z)
  \frac{\gamma}{\Gamma} + \hat{g}_c\left.\frac{1}{\rho}\frac{\p P}{\p
      r}\right|_0\right]\delta v_r  +
\frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right),\\
\ii \sigma Q &= c_s^2(r_0,z)\left(\ii k_r + \hat{g}_c
  \left.\frac{\p\ln{\rho}}{\p r}\right|_0\right)\delta v_r + c _s^2(r_0, z)\left( \left.\frac{\p\ln\rho}{\p
      z}\right|_0 \delta v_z + \frac{d\delta
    v_z}{dz}\right), \\
\ii \sigma \delta v_r & = \ii k_r W  -
\hat{g}_c\frac{1}{c_s^2(r_0,z)}\left.\frac{1}{\rho}\frac{\p P}{\p
  r}\right|_0Q - 2\Omega(r_0,z)\delta v_\phi,
\end{align}
where subscript $0$ denotes evaluation at $r=r_0$. To obtain the
linear problem in the VGSB, we take the radially-local 
approximation further and formally set $\hat{g}_c=0$, and define the
new co-ordinate $x=r - r_0$ with $k_r\to k_x$. It is clear that
this approximation improves with increasing $|k_x|$.

We may also reverse this procedure by reinstating the `global
correction' terms proportional to $\hat{g}_c$ in the linear VGSB
equations. Of course, doing so implies the basic state is not uniform
in $x$, which would be inconsistent with the full VGSB equations in
equilibrium.

@@

\section{Alternative inference of instability in the presence of weak
  vertical shear}\label{pert_theory}
We can also investigate the destabilizing effect of vertical shear
without explicitly solving the full governing equation,
Eq. \ref{iso_ode3}. We first
consider a system with $q\equiv0$, for which the eigenfunctions are
Hermite polynomials and the real eigenfrequencies are known. We then
perturb this system by introducing a weak vertical shear ($|q|\ll1$)
and linearize the governing equation. This procedure is
\begin{align}   
  q \to 0 + \delta q,\quad
  W \to \He_n + \delta W,\quad
  \hat{\sigma} \to \hat{\sigma} + \delta\hat{\sigma}. 
\end{align}
%where for this exercise we restore $\sigma$ as the eigenfrequency.

Linearizing the integral relation Eq. \ref{integral_relation1} in the
thin-disk limit, and taking the
imaginary part, we have
\begin{align}
  2\hat{\sigma}\imag(\delta\hat{\sigma})
  \left(1+\hat{k}^2\right) \int_{-\infty}^{\infty} w(\zhat)
  \He_n^2(\hat{z}) d\hat{z}
  = \delta q \epsilon \hat{k} 
  \int_{-\infty}^{\infty}
  w(\zhat)\hat{z}\He_n(\hat{z})\He_n^\prime(\zhat) d\hat{z}
\end{align}
Recognizing $\hat{z} = \He_1(\hat{z})$ and using $\He_n^\prime = n
\He_{n-1}$ we have
 \begin{align}
   2\hat{\sigma}\imag(\delta\hat{\sigma})
   \left(1+\hat{k}^2\right) \int_{-\infty}^{\infty} w(\zhat)
   \He_n^2(\hat{z}) d\hat{z}
   =\delta q \epsilon \hat{k} 
   \int_{-\infty}^{\infty}
   w(\zhat)\He_1(\hat{z})\He_n(\hat{z})n\He_{n-1}(\hat{z})d\hat{z}. 
 \end{align}

Finally, we evaluate the integrals using the results
\begin{align}
  \int_{-\infty}^{\infty}
  w(\xi) \He_k(\xi)\He_l(\xi) d\xi = \sqrt{2\pi}k!\delta_{kl} \quad
% \end{align}
\text{and} \quad
% \begin{align}
  \int_{-\infty}^{\infty}
  w(\xi) \He_m(\xi)\He_k(\xi)\He_l(\xi) d\xi =
  \frac{\sqrt{2\pi}m!k!l!}{(j-m)!(j-k)!(j-l)!}, 
\end{align}
where $j = (m+k+l)/2$. We then obtain
 \begin{align}
   2\hat{\sigma}\imag(\delta\hat{\sigma})
  \left(1+\hat{k}^2\right) = \delta q \epsilon \hat{k} n.
 \end{align}
Inserting the original eigenfrequency $\hat{\sigma} = \pm
\sqrt{n}(1+\hat{k}^2)^{-1/2}$ gives
\begin{align}
  \imag(\delta\hat{\sigma})= \pm \frac{1}{2}\delta q \epsilon
  \sqrt{n} \frac{\hat{k}}{\sqrt{1+\hat{k}^2}}. 
\end{align}
This result agrees with Eq. \ref{simple_growth} in the limit
$|q|\to0$. 
%for $n=1$, since the
%function $\He_1$ solves the governing equation exactly. 
For $\hat{k}\gg 1$ we have
\begin{align}
  \imag(\delta\hat{\sigma})\simeq \pm \frac{1}{2}\delta q \epsilon
  \sqrt{n} \sgn{\hat{k}}. 
\end{align}

>>>>

Using the notation and disk models in the main text, the
energy, density and $r$ momentum equations become 
\begin{align}
  \ii \sigma W  &=c _s^2(r_0, z)\left( \left.\frac{\p\ln\rho}{\p
      z}\right|_0 \delta v_z + \frac{\gamma}{\Gamma} \frac{d\delta
    v_z}{dz}\right) + \left[\ii k_rc_s^2(r_0,z)
  \frac{\gamma}{\Gamma} + \hat{g}_c\left.\frac{1}{\rho}\frac{\p P}{\p
      r}\right|_0\right]\delta v_r  +
\frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right),\label{gcorr_terms1}\\
\ii \sigma Q &= c_s^2(r_0,z)\left(\ii k_r + \hat{g}_c
  \left.\frac{\p\ln{\rho}}{\p r}\right|_0\right)\delta v_r + c _s^2(r_0, z)\left( \left.\frac{\p\ln\rho}{\p
      z}\right|_0 \delta v_z + \frac{d\delta
    v_z}{dz}\right),\label{gcorr_terms2} \\
\ii \sigma \delta v_r & = \ii k_r W  -
\hat{g}_c\frac{1}{c_s^2(r_0,z)}\left.\frac{1}{\rho}\frac{\p P}{\p
  r}\right|_0Q - 2\Omega(r_0,z)\delta v_\phi,\label{gcorr_terms3}
\end{align}
where subscript $0$ denotes evaluation at $r=r_0$. To obtain the
linear problem in the VGSB, we take the radially-local 
approximation further and formally set $\hat{g}_c=0$, $r\to x$ and
$k_r\to k_x$. It is clear that this approximation improves with
increasing $|k_x|$.

\subsection{Effect of global radial gradients}
We may also reverse the above procedure by reinstating the `global
correction' terms proportional to $\hat{g}_c$ in the linear VGSB
equations. Of course, doing so implies the basic state is not uniform
in $x$, which would be inconsistent with the full VGSB equations in
equilibrium. Nevertheless, we illustrate in Fig. \ref{gcorr_compare} the effect of
these global radial gradients on the growth rate of the fundamental
VSI.

The `global correction' terms do not change the qualitative dependence
of the fundamental VSI growth rates on the thermal relaxation
timescale. The difference is negligble for $\beta\lesssim 1$ (i.e.
fast thermal relaxation). The critical relaxation timescale for the
fundamental VSI is unaffected by these global radients.

However, in the adiabatic limit $\beta\to\infty$ the linear VGSB
equations give a positive growth rate as discussed in
\S\ref{analytic_adia}, which is inconsistent with the second
Solberg-Hoiland criterion (Eq. \ref{solberg2}). The correct behaviour
($\nu\to0$ as $\beta\to\infty$) is obtained with global correction.

It is interesting to note that there is a thermal relaxation timescale
that maximizes the mode decay rate. Here, it is $\beta\simeq2$ or
about $0.3$ orbital periods. This is, in fact, consistent with Fig. 12
of \cite{nelson13}.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/gcorr_compare} 
  \caption{Growth rate of
    the fundamental VSI mode as a function of the thermal
    relaxatin time $\beta$, computed from the linear VGSB equations used in
    the main text (dotted) and that including additional global
    radial gradient terms in  Eq. \ref{gcorr_terms1}---\ref{gcorr_terms3} (solid). 
    The disk parameters are $\Gamma=1.011$,
    $\gamma=1.4$ and $(p,q,\epsilon)=(-1.5,-1,0.05)$. The perturbation
    wavenumber is $\khat=20$. 
    {\bf waiting for
      high resolution} 
    \label{gcorr_compare}}  
\end{figure}

@@

% \section{Alternative inference of instability in the presence of weak
%   vertical shear}\label{pert_theory}
% We can also investigate the destabilizing effect of vertical shear
% without explicitly solving the full governing equation,
% Eq. \ref{iso_ode3}. We first
% consider a system with $q\equiv0$, for which the eigenfunctions are
% Hermite polynomials and the real eigenfrequencies are known. We then
% perturb this system by introducing a weak vertical shear ($|q|\ll1$)
% and linearize the governing equation. This procedure is
% \begin{align}   
%   q \to 0 + \delta q,\quad
%   W \to \He_n + \delta W,\quad
%   \hat{\sigma} \to \hat{\sigma} + \delta\hat{\sigma}. 
% \end{align}
% %where for this exercise we restore $\sigma$ as the eigenfrequency.

% Linearizing the integral relation Eq. \ref{integral_relation1} in the
% thin-disk limit, and taking the
% imaginary part, we have
% \begin{align}
%   2\hat{\sigma}\imag(\delta\hat{\sigma})
%   \left(1+\hat{k}^2\right) \int_{-\infty}^{\infty} w(\zhat)
%   \He_n^2(\hat{z}) d\hat{z}
%   = \delta q \epsilon \hat{k} 
%   \int_{-\infty}^{\infty}
%   w(\zhat)\hat{z}\He_n(\hat{z})\He_n^\prime(\zhat) d\hat{z}
% \end{align}
% Recognizing $\hat{z} = \He_1(\hat{z})$ and using $\He_n^\prime = n
% \He_{n-1}$ we have
%  \begin{align}
%    2\hat{\sigma}\imag(\delta\hat{\sigma})
%    \left(1+\hat{k}^2\right) \int_{-\infty}^{\infty} w(\zhat)
%    \He_n^2(\hat{z}) d\hat{z}
%    =\delta q \epsilon \hat{k} 
%    \int_{-\infty}^{\infty}
%    w(\zhat)\He_1(\hat{z})\He_n(\hat{z})n\He_{n-1}(\hat{z})d\hat{z}. 
%  \end{align}

% Finally, we evaluate the integrals using the results
% \begin{align}
%   \int_{-\infty}^{\infty}
%   w(\xi) \He_k(\xi)\He_l(\xi) d\xi = \sqrt{2\pi}k!\delta_{kl} \quad
% % \end{align}
% \text{and} \quad
% % \begin{align}
%   \int_{-\infty}^{\infty}
%   w(\xi) \He_m(\xi)\He_k(\xi)\He_l(\xi) d\xi =
%   \frac{\sqrt{2\pi}m!k!l!}{(j-m)!(j-k)!(j-l)!}, 
% \end{align}
% where $j = (m+k+l)/2$. We then obtain
%  \begin{align}
%    2\hat{\sigma}\imag(\delta\hat{\sigma})
%   \left(1+\hat{k}^2\right) = \delta q \epsilon \hat{k} n.
%  \end{align}
% Inserting the original eigenfrequency $\hat{\sigma} = \pm
% \sqrt{n}(1+\hat{k}^2)^{-1/2}$ gives
% \begin{align}
%   \imag(\delta\hat{\sigma})= \pm \frac{1}{2}\delta q \epsilon
%   \sqrt{n} \frac{\hat{k}}{\sqrt{1+\hat{k}^2}}. 
% \end{align}
% This result agrees with Eq. \ref{simple_growth} in the limit
% $|q|\to0$. 
% %for $n=1$, since the
% %function $\He_1$ solves the governing equation exactly. 
% For $\hat{k}\gg 1$ we have
% \begin{align}
%   \imag(\delta\hat{\sigma})\simeq \pm \frac{1}{2}\delta q \epsilon
%   \sqrt{n} \sgn{\hat{k}}. 
% \end{align}

### 2015-02-03 linear_adia.tex
updating

However, in the presence of vertical shear $q\neq0$,
Eq. \ref{adia_integral} suggests $\hat{\sigma}$ is complex for real
$\khat$, which implies the possibility of instability, regardless of
how much $\gamma$ deviates from unity. This is contradicts the second 
Solberg-Hoiland criterion (Eq. \ref{solberg2}) and is a result of the
radially-local approximation (Appendix \ref{global_corr}). To 
avoid this, we must restrict our attention to short thermal relaxation
timescales ($\beta\lesssim1$) and/or neutrally-stratified disks
($\gamma=\Gamma$) when allowing vertical shear.

>>>>

However, in the presence of vertical shear $q\neq0$,
Eq. \ref{adia_integral} suggests $\hat{\sigma}$ is complex for real
$\khat$, which implies the possibility of instability, regardless of
how much $\gamma$ deviates from unity. This is contradicts the second 
Solberg-Hoiland criterion (Eq. \ref{solberg2}) and is a result of the
radially-local approximation (Appendix \ref{global_corr}). To 
avoid this, we must restrict our attention to short thermal relaxation
timescales ($\beta\lesssim1$) and/or neutrally-stratified disks
($\gamma=\Gamma$) when $q\neq0$.
